---
title: "Mental Health Analysis"
format: html
editor: visual
---

### Step 1 Load the Packages & Data

-   **Research Question:** ***How do insurance and stigma-related barriers affect the likelihood of unmet mental health needs, and do these effects vary across gender and racial/ethnic groups?"***

```{r, include = FALSE}

# Load tidyverse for data manipulation and visualization
library(tidyverse)
library(dplyr)
```

```{r}
# Load the dataset from your local drive
df <- read.csv("C:/Users/srive/Downloads/SHAPEdata2022 (1).csv")
```

### Step 2 : Handle Missing Values in Key Variables

```{r}
# Define key variables required for your analysis
key_vars <- c("B8", "B9", "G1", "G12", "EthRace5_2022", "B10_4", "B10_5")

# View the number of missing values per key variable
na_counts <- colSums(is.na(df[key_vars]))
print(na_counts)  # Helps identify which variables need cleaning

# Create a clean dataset by removing rows with missing values in key variables
df_clean <- df %>%
  drop_na(all_of(key_vars))

# Confirm that missing values are removed
colSums(is.na(df_clean[key_vars]))
```

### STEP 3: Create Key Variables for Analysis

```{r}

df_clean <- df_clean %>%
  mutate(
    # Create binary outcome: 1 = unmet need, 0 = no unmet need
    unmet_need = ifelse(B8 == 1 & B9 == 2, 1, 0),
    
    # Recode gender: 1 = Male, 2 = Female, else = Other
    gender = factor(case_when(
      G1 == 1 ~ "Male",
      G1 == 2 ~ "Female",
      TRUE ~ "Other"
    )),
    
    # Recode race/ethnicity from numeric to descriptive labels
    race = factor(EthRace5_2022, levels = 1:5,
                  labels = c("Native", "Hispanic", "Asian", "Black", "White")),
    
    # Create binary variables for barriers
    insurance_barrier = ifelse(B10_5 == 1, 1, 0),
    stigma_barrier = ifelse(B10_4 == 1, 1, 0)
  )

```

# STEP 4: Examine Distribution of Unmet Needs by Gender and Race & **Barriers Across Gender and Race**

```{r}

# Create contingency tables
table(df_clean$unmet_need, df_clean$gender)
table(df_clean$unmet_need, df_clean$race)

# Perform Chi-square tests
chisq.test(table(df_clean$unmet_need, df_clean$gender))  # Gender vs Unmet Need
chisq.test(table(df_clean$unmet_need, df_clean$race))    # Race vs Unmet Need



# Stigma barrier differences by gender and race
chisq.test(table(df_clean$gender, df_clean$stigma_barrier))
chisq.test(table(df_clean$race, df_clean$stigma_barrier))

# Insurance barrier differences by gender and race
chisq.test(table(df_clean$gender, df_clean$insurance_barrier))
chisq.test(table(df_clean$race, df_clean$insurance_barrier))
```

# STEP 5: Unadjusted Logistic Regression Models

```{r}

# Unadjusted model for Gender
model_gender_unadj <- glm(unmet_need ~ gender, data = df_clean, family = binomial)
summary(model_gender_unadj)
exp(coef(model_gender_unadj))  # Odds Ratios (ORs)

# Unadjusted model for Race
model_race_unadj <- glm(unmet_need ~ race, data = df_clean, family = binomial)
summary(model_race_unadj)
exp(coef(model_race_unadj))

# Unadjusted model for Insurance Barrier
model_insurance_unadj <- glm(unmet_need ~ insurance_barrier, data = df_clean, family = binomial)
summary(model_insurance_unadj)
exp(coef(model_insurance_unadj))

# Unadjusted model for Stigma Barrier
model_stigma_unadj <- glm(unmet_need ~ stigma_barrier, data = df_clean, family = binomial)
summary(model_stigma_unadj)
exp(coef(model_stigma_unadj))
```

# **STEP 6: Fully Adjusted Logistic Regression Model**

```{r}

# Set "Female" as the reference group for gender
df_clean$gender <- relevel(df_clean$gender, ref = "Female")

# Set "White" as the reference group for race
df_clean$race <- relevel(df_clean$race, ref = "White")

#Logistic Regression Model : Insurance barrier

model_full1 <- glm(unmet_need ~ gender + race + insurance_barrier,
                  data = df_clean, family = binomial)

summary(model_full1)
exp(coef(model_full1))  # Adjusted Odds Ratios



#Logistic Regression Model : Stigma barrier

model_full2 <- glm(unmet_need ~ gender + race + stigma_barrier,
                  data = df_clean, family = binomial)

summary(model_full2)
exp(coef(model_full2))  # Adjusted Odds Ratios
```

# Step 7: Compare Adjusted and Unadjusted Effects ( 10% rule)

```{r}

#  Extract odds ratios from unadjusted models
or_insurance_unadj <- exp(coef(glm(unmet_need ~ insurance_barrier, data = df_clean, family = binomial)))
or_stigma_unadj    <- exp(coef(glm(unmet_need ~ stigma_barrier, data = df_clean, family = binomial)))

#  Extract odds ratios from adjusted models
or_insurance_adj <- exp(coef(model_full1))
or_stigma_adj    <- exp(coef(model_full2))

# Calculate % change in ORs for 10% rule
percent_change_insurance <- abs((or_insurance_unadj["insurance_barrier"] - or_insurance_adj["insurance_barrier"]) /
                                  or_insurance_unadj["insurance_barrier"]) * 100

percent_change_stigma <- abs((or_stigma_unadj["stigma_barrier"] - or_stigma_adj["stigma_barrier"]) /
                               or_stigma_unadj["stigma_barrier"]) * 100

# ðŸ–¨ï¸ Print results
print(paste("Percent change in OR for insurance_barrier:", round(percent_change_insurance, 2), "%"))
print(paste("Percent change in OR for stigma_barrier:", round(percent_change_stigma, 2), "%"))


```

# STEP 9: Interaction Models

```{r}


# Interaction: Gender Ã— Insurance Barrier
model_interact_gender_ins <- glm(unmet_need ~ gender + race + insurance_barrier +
                                   gender:insurance_barrier,
                                 data = df_clean, family = binomial)
summary(model_interact_gender_ins)

#  Interaction: Race Ã— Insurance Barrier
model_interact_race_ins <- glm(unmet_need ~ gender + race + insurance_barrier +
                                 race:insurance_barrier,
                               data = df_clean, family = binomial)
summary(model_interact_race_ins)

# Interaction: Gender Ã— Stigma Barrier
model_interact_gender_stigma <- glm(unmet_need ~ gender + race + stigma_barrier +
                                      gender:stigma_barrier,
                                    data = df_clean, family = binomial)
summary(model_interact_gender_stigma)

#  Interaction: Race Ã— Stigma Barrier
model_interact_race_stigma <- glm(unmet_need ~ gender + race + stigma_barrier +
                                    race:stigma_barrier,
                                  data = df_clean, family = binomial)
summary(model_interact_race_stigma)

```

# Step 10 : Bar Plot: Unmet Mental Health need by gender

```{r}

# Proportion table with percentage labels
gender_pct <- df_clean %>%
  count(gender, unmet_need) %>%
  group_by(gender) %>%
  mutate(perc = round(n / sum(n) * 100, 1))

# Bar plot with % labels
ggplot(gender_pct, aes(x = gender, y = perc, fill = unmet_need)) +
  geom_bar(stat = "identity", position = position_stack()) +
  geom_text(aes(label = paste0(perc, "%")),
            color= "white",
            position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Unmet Mental Health Need by Gender",
       x = "Gender", y = "Percentage",
       fill = "Unmet Need") +
  theme_minimal()


```

# Step 11: Bar Plot : Unmet Mental Need by Race

```{r}

# Proportion table with percentage labels
race_pct <- df_clean %>%
  count(race, unmet_need) %>%
  group_by(race) %>%
  mutate(perc = round(n / sum(n) * 100, 1))

# Bar plot with % labels
ggplot(race_pct, aes(x = race, y = perc, fill = unmet_need)) +
  geom_bar(stat = "identity", position = position_stack()) +
  geom_text(aes(label = paste0(perc, "%")),
            color ="white",
            position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Unmet Mental Health Need by Race",
       x = "Race", y = "Percentage",
       fill = "Unmet Need") +
  theme_minimal()


```
